Predicting Weak Lensing Statistics from Halo Mass Reconstructions

### 2015-08-21 doc/suli-report-swe.tex
Made final revisions for the Methods, Results, Discussion, and Conclusion sections of the SULI report. #81

%% Pangloss Cartoon
\begin{figure*}
    \centering
    \includegraphics[width=0.85\textwidth]{figs-swe/pangloss_cartoon.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{A cartoon model of how the \texttt{Pangloss} framework uses a dark matter halo model to reconstruct the foreground mass along the line of sight of a background source galaxy. A `lightcone' centered at a background source is constructed with radius $R$ and populated with all foreground galaxies contained in this volume. Each foreground object has an attached dark matter halo and physical distance to the line of sight (dashed line). The shear and convergence along the line of sight contributed by each halo is calculated using \cite{lensing_calc}, and the predicted ellipticity is determined using the sum of these contributions.}
    \label{pangloss_cartoon}
\end{figure*}

@@

Testing the first assumption is the main goal of this paper. The second assumption is reasonable as it is estimated that dark matter halos are 1-2 orders of magnitude more massive than the host galaxy \cite{smhr}. Clearly the third assumption will never be true for any observational data. However, this allows for a best-case scenario estimate of how well the \texttt{Pangloss} framework \textit{could} predict weak lensing effects given all possible information. This assumption can be relaxed by sampling a dark matter halo mass from an assumed stellar mass - halo mass relation. The fourth assumption is also unrealistic as most galaxies in sky surveys only have a less-reliable photometric redshift due to time constraints, but this again allows for a best-case estimate. This assumption could be relaxed by instead using photometric redshifts, adding random noise, and repeating the upcoming analysis on many realizations of the photometric redshifts.

@@

\subsection*{Generating Background Galaxies}
With a catalog of foreground galaxies chosen, a set of 3,240 background galaxies over a field of 324 arcmin$^2$ was generated, with density 10 per arcminute$^2$. As the The intrinsic orientation of each galaxy was sampled from a uniform distribution as, without lensing, there should be no preferred orientation due to the assumption of an isotropic universe. The magnitude of the galaxy ellipticities was sampled from a normal distribution with a standard deviation of 0.2, but any ellipticities with magnitude greater than one were re-sampled. Random ellipticity noise was sampled from a normal distribution with a standard deviation of 0.1 and added to the intrinsic ellipticities. Finally, each ellipticity was multiplied by $M=0.9$ to account for a multiplicative bias of 10\%.

@@

%% Ellipticity-Ellipticity Correlation
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/gg_corr.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The ellipticity-ellipticity correlation function for 3,240 sources at separation distances from 0.1 arcminutes to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. While the correlation of the intrinsic ellipticities before lensing (blue) shows no correlation, the correlation of the ray-traced ellipticities (green) and halo model predicted ellipticities (purple) both show significant positive correlation on scales up to 1 arcminute. The halo model predicts a small negative correlation above this scale, while the ray-tracing shows positive correlation on all scales.}
        %The correlation of the intrinsic ellipticities before lensing is plotted in blue, and shows no correlation. The correlation of the true ray-traced ellipticities is plotted in green and shows significant positive correlation on all scales. The halo predicted correlation, plotted in purple, shows significant positive correlation until a separation distance of 0.8 arcminutes.
        \label{gg_corr}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/gg_corr_series.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The ellipticity-ellipticity correlation functions for a series of halo model frameworks with increasing lightcone radii from $R=1$ arcminute to $R=8$ arcminutes, all plotted in purple with various line styles. These are compared to the same ray-traced correlation from Figure \ref{gg_corr}.}
        \label{gg_corr_series}
    \end{subfigure}
    \caption{}
\end{figure}

@@

Using the same catalog of background sources, this statistic was recalculated for the halo model ellipticities for various lightcone radii ranging from $R=1$ arcminute to $R=8$ arcminutes. The result is shown in Figure \ref{gg_corr_series}, where the series of model predicted correlation functions is compared to the ray-traced correlation function. As $R$ increases, the mean number of foreground galaxies contained in each lightcone grows quadratically from ${240\pm25}$ galaxies when $R=1$ arcminute to ${15,000\pm300}$ galaxies when $R=8$ arcminutes. Increasing the number of foreground objects, and thus increasing the mass and structure considered for shear and convergence predictions, systematically increased the predicted shear power on all scales. While the series of predicted correlation functions appear to be converging towards the ray-traced correlation function, there is clearly diminishing shear power gains as $R$ increases.

@@

The catalog of lensed ellipticities was also analyzed using the galaxy-mass correlation function with separation distances from 0.1 arcminutes to 2 arcminutes and a lightcone radius of 2 arcminutes. The result is shown in Figure \ref{ng_corr}. As with the ellipticity-ellipticity correlation function, there is no significant correlation for the intrinsic ellipticities at any scale as expected. The lensed ellipticities from the ray-tracing and halo model both have positive correlation on all calculated scales and are in agreement on scales from 0.2 arcminutes to 0.7 arcminutes. However, the halo model is still systematically underpredicting the shear power on scales above 0.7 arcminutes, although now only by an average of 23\%. A new feature is the large overprediction of shear power on separation scales smaller than 0.2 arcminutes. However, background sources this close to foreground halos are often in the strong lensing regime which is not currently accounted for by the \texttt{Pangloss} framework.

%% Galaxy-mass correlation function
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/ng_corr.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The galaxy-mass correlation function of 3,240 sources at separation distances from 0.1 arcminutes to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. There is no significant correlation of the intrinsic ellipticities (blue) as expected, and both the ray-traced correlation (green) and halo model correlation (purple) are positive on all scales. The green and purple curves are in agreement from 0.2 to 0.7 arcminutes, but have significant discrepancies on scales above and below this range.}
    \label{ng_corr}
\end{figure}

@@

A secondary goal of this research was to investigate how quickly the prediction for the shear and convergence for a single background source could be made. The mean CPU time required for the prediction per lightcone was ${80\pm10}$ ms for the first run. However, many small improvements to the coding infrastructure of \texttt{Pangloss} were made to speed up the prediction. The average CPU time for a lensing prediction per lightcone for four separate runs after various (and cumulative) changes is plotted in Figure \ref{cpu_plot}. Run 1 corresponds to the original \texttt{Pangloss} framework code, run 2 only wrote the needed quantities for this analysis to the catalog \texttt{Astropy} tables, run 3 changed the redshift grid binning to be done once for all halos in the drilling rather than for every individual halo in lensing, and run 4 optimized various lensing calculations with matrices whenever possible. Cumulatively, these changes alone accounted for a 70\% decrease in mean CPU time per lightcone and halved the initial variation.

%% Mean lightcone CPU time
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/profile_bar.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The mean CPU runtime per lightcone for various runs with different (and cumulative) code optimizations described in the text. The cumulative changes account for a 70\% decrease in the mean CPU time and a halving of the variation.}
    \label{cpu_plot}
\end{figure}

@@

While the current \texttt{Pangloss} framework can handle the lensing predictions of 3,240 in a 324 arcmin$^2$ field comfortably on a single processor, the goal is to scale up the framework to make predictions for background sources of the same number density in a 100 deg$^2$ field; this will require 3.6 million lightcones, each containing up to 1,000 foreground halos. Luckily, the prediction is trivially parallelizable as the shear and convergence calculation of each lightcone is independent of all other lightcones. This makes GPUs an attractive candidate for future work, as it would only take 360 GPUs with 10,000 threads each to carry out the prediction. Additionally, the 50 MB of RAM per lightcone is sufficiently small to fit 10,000 lightcones on a GPU.

@@

In this paper, a simple halo model was used to reconstruct all the mass along lines of sight in the Millennium Simulation to make predictions of how foreground mass weakly lensed the ellipticities of 3,240 generated background sources across a field of 324 arcmin$^2$. The lensed ellipticities were characterized globally using the ellipticity-ellipticity and galaxy-mass correlation functions and then compared with the same statistics from ray-traced data. There was a systematic-underprediction of shear power on all scales in the ellipticity-ellipticity correlation function as well as an absence of any predicted correlation on scales larger than 1 arcminute, suggesting missing features in the \texttt{Pangloss} framework. The galaxy-mass correlation function predicted by the halo model was in agreement with the ray-traced correlation function on scales from 0.2 to 0.7 arcminutes, but significantly overpredicted the shear power on smaller scales due to strong lensing effects. Now that a proof of concept has been demonstrated, there is work being done to scale up the \texttt{Pangloss} framework to make lensing predictions over 100 deg$^2$ using far more computational resources. While the results are encouraging for such a simple model, many strong and unrealistic assumptions were made in the methodology that limit its use in observational surveys. In the future, a similar analysis should be made after relaxing many of the assumptions to see how different the predictions by the halo model are from this best-case scenario.

>>>>

%% Pangloss Cartoon
\begin{figure*}
    \centering
    \includegraphics[width=0.85\textwidth]{figs-swe/pangloss_cartoon.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{A cartoon model of how the \texttt{Pangloss} framework uses a dark matter halo model to reconstruct the foreground mass along the line of sight of a background source galaxy. A `lightcone' centered at a background source is constructed with radius $R$ and populated with all foreground galaxies contained in this volume. Each foreground object has an attached dark matter halo and physical distance to the line of sight (dashed line). The shear and convergence along the line of sight contributed by each halo is calculated using \cite{lensing_calc}, and the predicted ellipticity is determined using the sum of these contributions and Equation \eqref{lensing}.}
    \label{pangloss_cartoon}
\end{figure*}

@@

Testing the first assumption is the main goal of this paper. The second assumption is reasonable as it is estimated that dark matter halos are on average 1-2 orders of magnitude more massive than the host galaxy \cite{smhr}. Clearly the third assumption will never be true for any observational data. However, this allows for a best-case scenario estimate of how well the \texttt{Pangloss} framework \textit{could} predict weak lensing effects given all possible information. This assumption can be relaxed by sampling a dark matter halo mass from an assumed stellar mass - halo mass relation. The fourth assumption is also unrealistic as most galaxies in sky surveys only have a less-reliable photometric redshift due to time constraints, but this again allows for a best-case estimate. This assumption could be relaxed by instead using photometric redshifts, adding random noise, and repeating the upcoming analysis on many realizations of the photometric redshifts.

@@

\subsection*{Generating Background Galaxies}
With a catalog of foreground galaxies chosen, a set of 3,240 background galaxies over a field of 324 arcmin$^2$ was generated, with density 10 per arcmin$^2$. The intrinsic orientation of each galaxy was sampled from a uniform distribution as, without lensing, there should be no preferred orientation due to the assumption of an isotropic universe. The magnitude of the galaxy ellipticities was sampled from a normal distribution with a standard deviation of 0.2, but any ellipticities with magnitude greater than one were re-sampled. Random ellipticity noise was sampled from a normal distribution with a standard deviation of 0.1 and added to the intrinsic ellipticities. Finally, each ellipticity was multiplied by $M=0.9$ to account for a multiplicative bias of 10\%.

@@

%% Ellipticity-Ellipticity Correlation
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/gg_corr.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The ellipticity-ellipticity correlation function for 3,240 sources at separation distances from 0.1 to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. While the correlation of the intrinsic ellipticities before lensing (blue) shows no correlation, the correlation of the ray-traced ellipticities (green) and halo model predicted ellipticities (purple) both show significant positive correlation on scales up to 1 arcminute. The halo model predicts a small negative correlation above this scale, while the ray-tracing shows positive correlation on all scales.}
        %The correlation of the intrinsic ellipticities before lensing is plotted in blue, and shows no correlation. The correlation of the true ray-traced ellipticities is plotted in green and shows significant positive correlation on all scales. The halo predicted correlation, plotted in purple, shows significant positive correlation until a separation distance of 0.8 arcminutes.
        \label{gg_corr}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/gg_corr_series.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The ellipticity-ellipticity correlation functions for a series of halo model frameworks with increasing lightcone radii from $R=1$ to $R=8$ arcminutes, all plotted in purple with various line styles. These are compared to the same ray-traced correlation from Figure \ref{gg_corr}.}
        \label{gg_corr_series}
    \end{subfigure}
    \caption{}
\end{figure}

@@

Using the same catalog of background sources, this statistic was recalculated for the halo model ellipticities for various lightcone radii ranging from $R=1$ to $R=8$ arcminutes. The result is shown in Figure \ref{gg_corr_series}, where the series of model predicted correlation functions is compared to the ray-traced correlation function. As $R$ increases, the mean number of foreground galaxies contained in each lightcone grows quadratically from ${240\pm25}$ galaxies when $R=1$ arcminute to ${15,000\pm300}$ galaxies when $R=8$ arcminutes. Increasing the number of foreground objects, and thus increasing the mass and structure considered for shear and convergence predictions, systematically increased the predicted shear power on all scales. While the series of predicted correlation functions appear to be converging towards the ray-traced correlation function, there is clearly diminishing shear power gains as $R$ increases.

@@

The catalog of lensed ellipticities was also analyzed using the galaxy-mass correlation function with separation distances from 0.1 to 2 arcminutes and a lightcone radius of 2 arcminutes. The result is shown in Figure \ref{ng_corr}. As with the ellipticity-ellipticity correlation function, there is no significant correlation for the intrinsic ellipticities at any scale as expected. The lensed ellipticities from the ray-tracing and halo model both show positive correlation on all calculated scales and are in agreement on separations from 0.2 to 0.7 arcminutes. However, the halo model is still systematically underpredicting the shear power on scales above 0.7 arcminutes, although now only by an average of 23\%. A new feature is the large overprediction of shear power on separation scales smaller than 0.2 arcminutes. However, background sources this close to foreground halos are often in the strong lensing regime which is not currently accounted for by the \texttt{Pangloss} framework.

%% Galaxy-mass correlation function
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/ng_corr.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The galaxy-mass correlation function of 3,240 sources at separation distances from 0.1 to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. There is no significant correlation of the intrinsic ellipticities (blue) as expected, and both the ray-traced correlation (green) and halo model correlation (purple) are positive on all scales. The green and purple curves are in agreement from 0.2 to 0.7 arcminutes, but have significant discrepancies on scales above and below this range.}
    \label{ng_corr}
\end{figure}

@@

A secondary goal of this research was to investigate how quickly the prediction for the shear and convergence for a single background source could be made. The mean CPU time required for the prediction per lightcone was ${80\pm10}$ ms for the first run. However, many small improvements to the coding infrastructure of \texttt{Pangloss} were made to speed up the prediction. The average CPU time for a lensing prediction per lightcone for four separate runs after various (and cumulative) changes is plotted in Figure \ref{cpu_plot}, with a final run time of $24\pm5$ ms. Run 1 corresponds to the original \texttt{Pangloss} framework code, run 2 only wrote the needed quantities for this analysis to the catalog \texttt{Astropy} tables, run 3 changed the redshift grid binning to be done once for all halos in the drilling rather than for every individual halo in lensing, and run 4 optimized various lensing calculations with matrices whenever possible. Cumulatively, these changes alone accounted for a 70\% decrease in mean CPU time per lightcone and halved the initial variation.

%% Mean lightcone CPU time
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/profile_bar.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The mean CPU runtime per lightcone for various runs with different (and cumulative) code optimizations described in the text. The cumulative changes account for a 70\% decrease in the mean CPU time and a halving of the variation. Currently, the prediction can be made for a single lightcone in\\$24\pm5$ ms.}
    \label{cpu_plot}
\end{figure}

@@

While the current \texttt{Pangloss} framework can handle the lensing predictions of 3,240 sources in a 324 arcmin$^2$ field comfortably on a single processor, the goal is to scale up the framework to make predictions for background sources of the same number density in a 100 deg$^2$ field; this will require 3.6 million lightcones, each containing up to 1,000 foreground halos. Luckily, the prediction is trivially parallelizable as the shear and convergence calculation of each lightcone is independent of all other lightcones. This makes GPUs an attractive candidate for future work, as it would only take 360 GPUs with 10,000 threads each to carry out the prediction. Additionally, the 50 MB of RAM per lightcone is sufficiently small to fit 10,000 lightcones on a GPU.

@@

In this paper, a simple halo model was used to reconstruct all the mass along lines of sight in the Millennium Simulation to make predictions of how foreground mass weakly lensed the ellipticities of 3,240 generated background sources across a field of 324 arcmin$^2$. The lensed ellipticities were characterized globally using the ellipticity-ellipticity and galaxy-mass correlation functions and then compared with the same statistics from ray-traced data. There was a systematic-underprediction of shear power on all scales in the ellipticity-ellipticity correlation function as well as an absence of any predicted correlation on scales larger than 1 arcminute, suggesting missing features in the \texttt{Pangloss} framework. The galaxy-mass correlation function predicted by the halo model was in agreement with the ray-traced correlation function on scales from 0.2 to 0.7 arcminutes, but significantly overpredicted the shear power on smaller scales due to strong lensing effects. Now that a proof of concept has been demonstrated, there is work being done to scale up the \texttt{Pangloss} framework to make lensing predictions over 100 deg$^2$ using far more computational resources. While the results are encouraging for such a simple model, many strong and unrealistic assumptions were made in the methodology that limit its use in observational surveys. In the future, a similar analysis should be made after relaxing some of the assumptions to see how different the predictions by the halo model are from this best-case scenario.
